---
title: "Homework solutions"
subtitle: "Environmental Statistics 1"
author: "DAG"
format:  pdf
number-sections: true
toc: false
editor: source
geometry:
  - top=20mm
  - left=20mm
  - bottom=20mm
  - right=20mm

monofontoptions: 
  - Scale=0.8
---

```{r}
#| echo: false
library(knitr)
opts_chunk$set(fig.path='figure/', fig.align='center', fig.width=4, fig.height=4.3, fig.show='hold', cache=TRUE, tidy=F, tidy.opts=list(width.cutoff=60), size="small")
```


## Bernoulli regression

*In the 1996 presidential elections in the USA, Bill Clinton ran against Bob Dole. faraway’s nes96 are the results
of a poll, along with a range of predictors.*

### Load the data and look at the summary of it.


```{r}
library("faraway")
library("ggplot2")
data("nes96")
```


### Plot vote as a function of population density.


```{r}

plot(vote ~ log(popul+1), data=nes96)

```

### Fit a GLM to represent this relationship


```{r}

fm <- glm(vote ~ log(popul+1), data=nes96, family=binomial)
summary(fm)

```
*Yes it is: the response variable is internally coded as Clinton=0, Dole=1 (in alphabetical sequence). Hence the slope
estimate is of the probability of 1 (i.e. voting Dole) as affected by population density.*












